A Flexible Machine Learning-Aware Architecture\\ for Future WLANs

### 2020-01-09 LaTeX files/flexible_ml_architecture_wlans.tex
Based on reviewers’ comments, we have applied some changes to the article (to be continued).

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
Lots of hopes have been placed in Machine Learning (ML) as a key enabler of future wireless networks. By taking advantage of the large volumes of data generated by networks, ML is expected to deal with the ever-increasing complexity of networking problems. Unfortunately, current networking systems are not yet prepared for supporting the ensuing requirements of ML-based applications, especially for enabling procedures related to data collection, processing, and output distribution. This article points out the architectural requirements that are needed to pervasively include ML as part of future wireless networks operation. To this aim, we propose to adopt the International Telecommunications Union (ITU) unified architecture for 5G and beyond. Specifically, we look into Wireless Local Area Networks (WLANs), which, due to their nature, can be found in multiple forms, ranging from cloud-based to edge-computing-like deployments. Based on the ITU's architecture, we provide insights on the main requirements and the major challenges of introducing ML to the multiple modalities of WLANs.
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Machine Learning, Future Networks, Wireless Local Area Networks,
Architecture, ITU
\end{IEEEkeywords}

@@

\begin{table*}[t!]
	\caption{Machine learning methods, algorithms, potential networking applications, and examples of input data.}
	\label{table:ml_taxonomy}
	\centering
	\begin{tabular}{|p{.09\textwidth}|p{.28\textwidth}|p{.25\textwidth}|p{.25\textwidth}|}
		\hline
		\textbf{ML method} & \textbf{Algorithms} & \textbf{Potential applications} & \textbf{Examples of input data} \\\hline
		Supervised Learning & Linear classifiers, Regression methods (e.g., ARIMA), Neural Networks, Hidden Markov Models, Random Forest, Support Vector Machines, k-Nearest Neighbors, Principal Component Analysis &Traffic forecasting, mobility pattern prediction, flow classification, routing, anomaly detection, spectrum management, failure detection, QoE prediction & IP traffic matrices, temporal user location, availability of routing paths, application data, channel measurements, performance metrics \\\hline
		Unsupervised Learning & Clustering, Mixture models, Generative models, Non-Negative Matrix Factorization, Evolutionary algorithms & Traffic classification, virtual topology design, path computation, intruder detection, signal separation& IP traffic matrices, historical end-to-end bit-rate, received symbols \\\hline
		Reinforcement Learning & Monte Carlo, Q-learning, SARSA, Deep Q Network, Actor-critic, Multi-Armed Bandits, Learning automaton & Power control, rate adaptation, routing, channel selection, spatial reuse, smart caching, traffic offloading, cognitive channel access, energy harvesting& Channel measurements, link status, performance metrics (e.g., throughput, delay) \\\hline
	\end{tabular}
\end{table*}

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by the IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
Wireless communications have reached a point where a paradigm shift is required for satisfying the increasing needs of next-generation applications \cite{osseiran2014scenarios}. Based on the current trend, Artificial Intelligence (AI), and more precisely Machine Learning (ML), are expected to conduct a revolution, especially regarding the network planning, operation, and management of the fifth and sixth generation (5G/6G) of mobile communications.

ML is meant to empower a computational system for learning automatically, based on experience, so that future situations can be properly managed without having been programmed explicitly. Concerning wireless communications, there is a huge amount of unexploited data generated at both infrastructure and user levels, which could be extremely useful to ML mechanisms for learning complex patterns. According to the information extracted by these methods, network performance can be improved. For instance, the behavior of users in a given network-oriented service can be predicted through ML, based on the information from previous sessions. This information would enable to better accommodate network resources in future sessions, thus leading to an optimized solution.

Unfortunately, the potential benefits of ML solutions for real networks are currently limited by the existing infrastructure, which is not yet prepared to accommodate ML-oriented tasks such as data collection, processing, and output distribution. Instead, current networking systems are typically meant for delivering content, without taking into account the underlying characteristics of the processes that generate it.

The first steps towards AI-enabled networking are currently being made in 5G through network function virtualization (NFV). Unlike traditional hardware-based networks, NFV allows rapid elasticity and fast reconfiguration on assigning network resources, which is particularly useful to enable verticals such as autonomous driving in the automotive sector or smart manufacturing in Industry 4.0. Besides, network virtualization is useful to boost inter-operator coordination, which would bring the ML operation to a macro-scale level, thus counting with vast information and computation resources.

@@

To contribute to the evolution of wireless communications towards AI-based systems, we provide a realization of an ML-aware architecture for IEEE 802.11 Wireless Local Area Networks (WLANs), which will be an essential part of the 5G/6G ecosystem in the unlicensed bands. Unlike for cellular networks, WLANs have received much less attention when designing AI-aware architectural solutions. The fact is that cellular-based deployments fit in perfectly with big data analytics, due to the vast amount of data and high computation resources available for mobile network operators (MNOs). In contrast, WLANs pose a set of specific challenges resulting from their multiple deployment modes (e.g., campus network, residential usage) and their typical decentralized nature. Despite WLANs can count with plenty of data to be used by ML methods in large and planned deployments, we find other decentralized scenarios that lack powerful centralized equipment. In the latter cases, huge computing and processing operations cannot be endowed to the ML operation.

@@

\subsection{Machine Learning in Communications}
% 5G and ML as enabler
The proliferation of new communication-based applications is defining the shape of future networks through a set of strict requirements \cite{itu2019use}. Some examples are Vehicle to Everything (V2X), Industry 4.0, and Virtual Reality / Augmented Reality (VR/AR). Those applications are really challenging in terms of bandwidth, latency, reliability, and scalability. In this regard, the main pillars to be fulfilled by next-generation wireless networks are:
\begin{enumerate}
	\item Provide a very high data rate (10-20 Gbps).
	\item Support a massive amount of connected devices (1,000,000 devices/km$^2$).
	\item Dramatically reduce the latency ($<5$ ms) whilst maintaining a low packet error rate (1 packet lost for every 10$^5$ packets sent).
\end{enumerate}

In 5G, the previous concepts are referred to as enhanced mobile broadband (eMBB), massive machine to machine (M2M) and Internet of Things (IoT) communication (mMTC), and ultra-reliable and low latency communication (uRLLC), respectively. Similarly, IEEE 802.11 groups are also considering those aspects in the design of next-generation amendments, such as high efficiency (HE) IEEE 802.11ax and extreme high throughput (EHT) IEEE 802.11be.

@@

The actual utility of ML lies in those problems that are hard to solve by hand-programming due to their underlying complex patterns (e.g., network traffic prediction). Formally, a machine is said to learn if it improves the performance $\mathbb{P}$ obtained from undertaking task $\mathbb{T}$, based on the gathered experience $\mathbb{E}$ \cite{mitchell1997machine}. Different ML techniques have been categorized in multiple ways, but the most common taxonomy differentiates between Supervised Learning (labeled data is used for training), Unsupervised Learning (no labels are used on input data), and Reinforcement Learning (exploration-exploitation trade-off with label/unlabeled data). Table \ref{table:ml_taxonomy} provides a list of algorithms and potential networking applications for each type of ML technique, as well as some examples of input data for these methods. For further details, we address the interested reader to \cite{jiang2016machine, zhang2019deep, usama2019unsupervised}, which survey a plethora of ML-based applications for networking.

% 1 - Architectural proposals	
Apart from the specific ML solutions to problems in communications, some efforts have been made towards enabling AI-aware networking in more general terms. In this regard, several architectural proposals have been provided so far \cite{bi2015wireless,chih2017big,wang2018machine}. Most of the referenced works agree in the necessary steps for enabling big data analytics in cellular deployments, which roughly are: (1) data collection, (2) data preparation, (3) data analysis, and (4) decision making. Nevertheless, none of these works provide architectural guidelines to introduce ML to wireless networks. In contrast, the ITU's architecture looks deeper into the ML operation and targets the actual procedures involving information gathering, processing, and communication. In addition to the architectural guidelines, the ITU-T has provided a data handling framework for ML-aware networks \cite{itu2019data}, which defines processes concerning data collection, processing, and output distribution.

\subsection{Machine Learning-Enabled Use Cases in IEEE 802.11 WLANs}
To showcase the potential of applying AI in IEEE 802.11 WLANs, we next describe a set of use cases where ML allows improving the network operation and performance.

\subsubsection{OFDMA-Based Smart Network Slicing} 
Network slicing is one of the hottest topics in 5G because it allows separating network resources virtually to meet diverse application requirements. In next-generation WLANs, network slicing can be realized through the allocation of radio resources via orthogonal frequency-division multiple access (OFDMA). However, the heterogeneity of applications running in multiple devices and their subsequent elasticity prevent to assign frequency resources easily.

The application of ML can overcome the aforementioned challenges by making predictions on the user requirements so that the access network can be optimized. As an example, Fig. \ref{fig:use_cases} shows a scenario where multiple users have different requirements, based on the applications they use. While the central controller can make predictions on user behavior, the local schedulers can consider information such as the user profile, the current performance, and the environmental circumstances. In this regard, the Access Point (AP) can allocate the most suitable OFDMA resources to each device, according to its predicted needs and current status of the network.

@@

\subsubsection{Cloud-Based User Association and Handover}
Most of the current user association and handover procedures held in WLANs typically rely on strongest signal first (SSF) mechanisms. This might be problematic in terms of load balancing, thus potentially leading to severe performance degradation at saturated APs. By introducing ML, it is possible to handle contextual information such as the traffic load at the APs, which can support decision-making. Furthermore, mobility pattern prediction and predicted user requirements can be included in the system, thus empowering the association and handover mechanisms with insightful information. %\textcolor{orange}{Predictions can be provided, for instance, by Deep Neural Networks (DNN).}

@@

\subsubsection{Inference-Based Coordinated Scheduling}

Contrary to traditional cellular-type networks, WLAN deployments can be chaotic, especially in residential scenarios where anyone can set-up an AP and create a wireless network. This typically leads to scenarios with complex interactions among different Basic Service Sets (BSSs), which prevent the existing scheduling approaches to ensuring proper quality of service (QoS). Nevertheless, ML can be used to infer these interactions and provide a solution accordingly. In particular, through coordinated ML-assisted scheduling, different APs can trigger uplink/downlink transmissions from/to the appropriate stations (STAs), thus increasing the network throughput and reducing the number of packet collisions.

\subsubsection{Reinforcement Learning-Based Spatial Reuse} 
In contrast to centralized and cooperative ML approaches, local optimization is required for independent WLANs. In this regard, Spatial Reuse (SR) aims to improve channel utilization by means of sensitivity adjustment mechanisms. However, selecting the best sensitivity threshold is not trivial, given the complex spatial interactions that occur among WLANs and the high action space. With this purpose, RL can be introduced to improve spectral efficiency in a decentralized manner, so that each WLAN learns based on experience.

@@

\begin{figure*}[!ht]
	\centering
	\includegraphics[width=0.5\textwidth]{itu_ml_architecture}
	\caption{ITU logical architecture for future networks \cite{itu2019architecture}.}
	\label{fig:itu_ml_architecture}
\end{figure*}

@@

Figure \ref{fig:itu_ml_architecture} shows the elements that compose the ML overlay (management subsystem, multi-level ML Pipeline, and closed-loop subsystem), which are further described in the following subsections. Based on this overlay, ML applications can be instantiated in the logical entities (represented by white boxes), which are managed in a standard manner. %Communication among entities is done through input and output interfaces.

\subsection{Management Subsystem} 
The management subsystem is in charge of the deployment and the orchestration of the ML services so that they can operate in the underlying network. To that purpose, the Machine Learning Function Orchestrator (MLFO) entity is defined. The MLFO is first instantiated by an intent, which is a declarative mechanism that provides details on the ML use case to be applied. The ML intent, therefore, defines the methods and policies to be applied throughout the network and instantiates the elements of the multi-level ML Pipeline in the corresponding devices. In particular, the intent can use a meta language to define the policies and constraints that characterize the use case.

\subsection{Multi-Level Machine Learning Pipeline} 
The multi-level ML Pipeline is formed by a set of logical entities that perform the actual ML activity in a given network underlay. The ML Pipeline is instantiated and managed by the management subsystem and it is mainly in charge of the data collection and pre-processing, ML model application, and output distribution. The logical entities that compose the ML Pipeline are as follows:
\begin{itemize}
	\item \textbf{Source (src):} element that generates data to be used by the ML mechanism.
	\item \textbf{Collector (C):} element responsible for collecting the data generated by source nodes.
	\item \textbf{Pre-processor (PP):} element that prepares the data collected by the collector for its utilization by the ML mechanism.
	\item \textbf{Model (M):} ML model that is applied according to the use case.
	\item \textbf{Policy (P):} provides a set of constraints and/or guidelines that delimit the behavior of the model.
	\item \textbf{Distributor (D):} element that spreads the ML output across all the corresponding targets (or sinks).
	\item \textbf{Sink (sink):} target of the ML output that is allocated by the distributor.
\end{itemize}

\subsection{Closed-loop Subsystem} 
In order to address network dynamics, the ML operation is assisted by a closed-loop subsystem. In particular, a sandbox environment can be used to reproduce (internal network) or simulate (simulator/emulator) the effect of certain ML optimization, so that additional information is provided to the system beforehand. As for the ML Pipeline, the closed-loop subsystem is orchestrated by the management subsystem. Network simulators (e.g., ns-3, Komondor \cite{barrachina2019komondor}) are examples of closed-loop subsystems, which can serve two purposes: \emph{i)} generate synthetic data to be used for training, and \emph{ii)} conducting a simulation for devising the potential of a given ML method to be applied afterward on the real network.

@@

IEEE 802.11 WLANs are versatile in the sense that many different kinds of deployments can be realized. Roughly, we differentiate between two main types:
\begin{itemize}
	\item \textbf{Enterprise:} a set of WLANs can be jointly operated from the edge and/or the cloud, thus providing management and orchestration functionalities (e.g., centralized authentication, channel allocation). Enterprise-like deployments are realized through extended service sets (ESS) and can be typically found in environments controlled by a single network operator, such as university campuses, offices, stadiums, etc. 
	\item \textbf{Residential:} each WLAN is responsible for its own management and operation. In the context of residential scenarios (but not limited to), ad-hoc WLANs such as personal hotspots are gaining popularity for infrastructureless communications (i.e., peer-to-peer). Ad-hoc WLANs are realized through the independent basic service set (IBSS).
\end{itemize}

Figure \ref{fig:overview_learning_approaches} illustrates the enterprise and residential-like deployments, as well as a set of IEEE 802.11 mechanisms (summarized in Table \ref{tab:opportunities_amendments}) that can potentially enable the utilization of ML in WLANs. These mechanisms can be used for data collection, training, and output distribution. By taking advantage of the current IEEE 802.11 mechanisms, ML can be introduced to WLANs in a non-intrusive manner, thus facilitating its immediate adoption.

@@

@@

%%% CHALLENGES IN IEEE 802.11 WLANs
\subsection{Challenges in IEEE 802.11 WLANs}
\label{section:ieee_80211_wlans}
The application of ML methods in WLANs is tightly tied to the technological challenges posed by these types of networks. The major challenges encountered in wireless communications stand for fast data expiry and lack of resources for data handling (e.g., storage, computation, information exchange). Regarding Wi-Fi networks, we find the following challenges:
\begin{enumerate}
	\item \textbf{Network dynamics:} channel fluctuations (e.g., due to multipath fading), users mobility and varying traffic needs entail a big challenge to ML applications. As a result of network dynamics, ML methods need to constantly adapt to the environment, which may be achieved through continuous re-training.
	\item \textbf{Limited communication resources:} since Wi-Fi operates under unlicensed bands, resources are scarce and shared. Thus, any potential communication required by a certain ML mechanism (as for distributed learning) may fail or be delayed if the medium is congested. As a result, the ML operation must be robust and resilient enough in order to react to potential communication issues.
	\item \textbf{Limited computation and storage resources:} computation and storage resources may also be limited in WLANs, especially in residential-like deployments. Therefore, the ML operation should include computation-efficient procedures. Another implication of limited resources lies in the availability of information to be used by ML algorithms, especially for online learning methods.
	\item \textbf{Adversarial environment:} in many cases, Wi-Fi deployments are chaotic in the sense that many overlapping WLANs coexist without any kind of cooperation (e.g., residential buildings). This is a particularly interesting challenge for ML operation, which may lead to adversarial settings where different agents compete for the same resources. Moreover, multi-vendor devices may implement different ML mechanisms that may lead to clashing interests.
	\item \textbf{Legacy devices:} in a similar way to the previous case, ML-empowered WLANs may coexist with other legacy networks that do not perform any intelligent operation. It is then required for ML methods to be aware of those legacy devices, so that unfairness situations are avoided.
\end{enumerate}

>>>>

\usepackage{color}
\usepackage[dvipsnames]{xcolor}
\usepackage{soul}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
Lots of hopes have been placed \textcolor{orange}{on} \textcolor{blue}{machine learning (ML)} as a key enabler of future wireless networks. By taking advantage of the large volumes of data generated by networks, ML is expected to deal with the ever-increasing complexity of networking problems. Unfortunately, current \textcolor{orange}{networks} are not yet prepared \textcolor{orange}{to support} the ensuing requirements of ML-based applications \textcolor{orange}{in terms of} data collection, processing, and output distribution. This article points out the architectural requirements that are needed to pervasively include ML as part of future wireless networks operation. \textcolor{orange}{Specifically, we look into \textcolor{blue}{wireless local area networks (WLANs)}, which, due to their nature, can be found in multiple forms, ranging from cloud-based to edge-computing-like deployments. In particular,} we propose to adopt the International Telecommunications Union (ITU) unified architecture for 5G and beyond. Based on the ITU's architecture, we provide insights on the main requirements and the major challenges of introducing ML to the multiple modalities of WLANs. %\textcolor{blue}{Finally, numerical results show the superiority of ML-based methods enabled by the architecture in front of currently adopted approaches.}
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
\textcolor{blue}{Architecture, Future Networks, ITU, Machine Learning, Wireless Local Area Networks}
\end{IEEEkeywords}

@@

\begin{table*}[t!]
	\caption{Machine learning methods, algorithms, potential networking applications, and examples of input data.}
	\label{table:ml_taxonomy}
	\centering
	\begin{tabular}{|p{.09\textwidth}|p{.28\textwidth}|p{.25\textwidth}|p{.25\textwidth}|}
		\hline
		\textbf{ML method} & \textbf{Algorithms} & \textbf{Potential applications} & \textbf{Examples of input data} \\\hline
		Supervised learning & Linear classifiers, regression methods (e.g., \textcolor{blue}{autoregressive integrated moving average}), neural networks, hidden Markov models, random forest, support vector machines, k-nearest neighbors, principal component analysis &Traffic forecasting, mobility pattern prediction, flow classification, routing, anomaly detection, spectrum management, failure detection, QoE prediction & IP traffic matrices, temporal user location, availability of routing paths, application data, channel measurements, performance metrics \\\hline
		Unsupervised learning & Clustering, mixture models, generative models, non-negative matrix factorization, evolutionary algorithms & traffic classification, virtual topology design, path computation, intruder detection, signal separation& IP traffic matrices, historical end-to-end bit-rate, received symbols \\\hline
		Reinforcement learning & Monte Carlo, Q-learning, \textcolor{blue}{state-action-reward-state-action (SARSA)}, deep Q network, actor-critic, multi-armed bandits, learning automaton, \textcolor{blue}{Markov decision processes} & Power control, rate adaptation, routing, channel selection, spatial reuse, smart caching, traffic offloading, cognitive channel access, energy harvesting, \textcolor{blue}{energy efficiency} & Channel measurements, link status, performance metrics (e.g., throughput, delay), \textcolor{blue}{server occupation, power consumption} \\\hline
	\end{tabular}
\end{table*}

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by the IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
Wireless communications have reached a point where a paradigm shift is required for satisfying the increasing needs of next-generation applications \cite{osseiran2014scenarios}. Based on the current trend, \textcolor{blue}{artificial intelligence (AI)}, and more precisely \textcolor{blue}{machine learning (ML)}, are expected to conduct a revolution, especially regarding the network planning, operation, and management of the 5th and 6th generation (5G/6G) of mobile communications.

ML is meant to empower a computational system for learning automatically, based on experience, so that future situations can be properly managed without having been programmed explicitly. Concerning wireless communications, there is a huge amount of unexploited data generated at both infrastructure and user levels, which could be extremely useful \textcolor{orange}{\st{to ML mechanisms}} for learning complex patterns\textcolor{orange}{, which may allow to improve network performance}. For instance, the behavior of users in a given network-oriented service can be predicted through ML, based on the information from previous sessions. \textcolor{orange}{Based on these predictions, network resources can be properly accommodated} in future sessions, thus leading to an optimized solution.

Unfortunately, the potential benefits of ML \textcolor{orange}{\st{solutions}} for real networks are currently limited by the existing infrastructure, which is not yet prepared to accommodate ML-oriented tasks such as data collection, processing, and output distribution. Instead, current networking systems are typically meant for delivering content, without taking into account the underlying characteristics of the processes that generate it.

The first steps towards AI-enabled networking are currently being made in 5G through network function virtualization (NFV). Unlike traditional hardware-based networks, NFV allows rapid elasticity and fast reconfiguration on assigning network resources, which is particularly useful to enable verticals such as autonomous driving in the automotive sector or smart manufacturing in \textcolor{blue}{industry 4.0}. Besides, network virtualization is useful to boost inter-operator coordination, which would bring the ML operation to a macro-scale level, thus counting with vast information and computation resources.

@@

To contribute to the evolution of wireless communications towards AI-based systems, we provide a realization of an ML-aware architecture for IEEE 802.11 \textcolor{blue}{wireless local area networks (WLANs)}, which will be an essential part of the 5G/6G ecosystem in the unlicensed bands. Unlike for cellular networks, WLANs have received much less attention when designing AI-aware architectural solutions. The fact is that cellular-based deployments fit in perfectly with big data analytics, due to the vast amount of data and high computation resources available for mobile network operators (MNOs). In contrast, WLANs pose a set of specific challenges resulting from their multiple deployment modes (e.g., campus network, residential usage) and their typical decentralized nature. Despite WLANs can count with plenty of data to be used by ML methods in large and planned deployments, we find other  \textcolor{orange}{residential-type} scenarios that lack \textcolor{orange}{of} powerful centralized equipment. In  \textcolor{orange}{these} cases, huge computing and processing  \textcolor{orange}{resources} cannot be endowed to the ML operation.

@@

\subsection{Machine Learning in Communications}
% 5G and ML as enabler
The proliferation of new communication-based applications is defining the shape of future networks through a set of strict requirements \cite{itu2019use}. Some examples are \textcolor{blue}{vehicle to everything (V2X)}, \textcolor{blue}{industry 4.0}, and \textcolor{blue}{virtual reality / augmented reality (VR/AR)}.  \textcolor{orange}{These} applications are really challenging in terms of bandwidth, latency, reliability, and scalability. In this regard, next-generation wireless networks  \textcolor{orange}{are expected to}:
\begin{enumerate}
	\item Provide a very high data rate (10-20 Gbps).
	\item Support a massive amount of connected devices (1,000,000 devices/km$^2$).
	\item Dramatically reduce the latency ($<5$ ms) whilst maintaining a low packet error rate (1 packet lost for every 10$^5$ packets sent).
\end{enumerate}

In 5G, the previous concepts are referred to as enhanced mobile broadband (eMBB), massive machine to machine and \textcolor{blue}{Internet of things (IoT)} communication (mMTC), and ultra-reliable and low latency communication (uRLLC), respectively. Similarly, IEEE 802.11 groups are also considering those aspects in the design of next-generation amendments, such as high efficiency (HE) IEEE 802.11ax and extreme high throughput (EHT) IEEE 802.11be.

@@

The actual utility of ML lies in those problems that are hard to solve by hand-programming due to their underlying complex patterns (e.g., network traffic prediction). Formally, a machine is said to learn if it improves the performance $\mathbb{P}$ obtained from undertaking task $\mathbb{T}$, based on the gathered experience $\mathbb{E}$ \cite{mitchell1997machine}. Different ML techniques have been categorized in multiple ways, but the most common taxonomy differentiates between supervised learning (labeled data is used for training), unsupervised learning (no labels are used on input data), and reinforcement learning (exploration-exploitation trade-off with label/unlabeled data). Table \ref{table:ml_taxonomy} provides a list of algorithms and potential networking applications for each type of ML \textcolor{blue}{techniques}, as well as some examples of input data  \textcolor{orange}{to be used by these} methods. For further details, we address the interested reader to \cite{jiang2016machine, zhang2019deep, usama2019unsupervised}, which survey a plethora of ML-based applications for networking.

% 1 - Architectural proposals	
Apart from the specific ML solutions to problems in communications, some efforts have been made towards enabling AI-aware networking in more general terms. In this regard, several architectural proposals have been provided so far \cite{bi2015wireless,chih2017big,wang2018machine}. Most of the referenced works agree in the necessary steps for enabling big data analytics in cellular deployments\textcolor{orange}{\st{, which roughly are}}: (1) data collection, (2) data preparation, (3) data analysis, and (4) decision making. Nevertheless, none of these works provide architectural guidelines to introduce ML to wireless networks. In contrast, the ITU's architecture looks deeper into the ML operation and targets the actual procedures involving information gathering, processing, and communication. In addition to the architectural guidelines, the ITU-T has provided a data handling framework for ML-aware networks \cite{itu2019data}, which defines processes concerning data collection, processing, and output distribution.

\subsection{Machine Learning-Enabled Use Cases in WLANs}
To showcase the potential of applying AI in IEEE 802.11 WLANs, we next describe a set of use cases where ML allows improving the network operation and performance.

\subsubsection{OFDMA-Based Smart Network Slicing} 
Network slicing is one of the hottest topics in 5G because it allows separating network resources virtually to meet diverse application requirements. In next-generation WLANs, network slicing can be realized through the allocation of radio resources via orthogonal frequency-division multiple access (OFDMA). However, the heterogeneity of applications running in multiple devices and their subsequent elasticity prevent \textcolor{orange}{allocating} frequency resources easily. \textcolor{orange}{To solve this, ML can be used to make} predictions on the user requirements so that the access network can be optimized.

As an example, Fig. \ref{fig:use_cases} shows a scenario where multiple users \textcolor{orange}{operate under} different requirements, based on the applications they use. While the central controller can make predictions on user behavior, the local schedulers can consider information such as the user profile, the current performance, and the environmental circumstances. In this regard, the \textcolor{blue}{access point (AP)} can allocate the most suitable OFDMA resources to each device, \textcolor{orange}{based on the} predicted needs and \textcolor{orange}{network} status.

@@

\subsubsection{Cloud-Based User Association and Handover}
Most of the current user association and handover procedures held in WLANs typically rely on strongest signal first (SSF) mechanisms. This might be problematic in terms of load balancing \textcolor{orange}{and can potentially lead} to severe performance degradation \textcolor{orange}{in} saturated \textcolor{blue}{basic service sets (BSSs)}. By introducing ML, it is possible to handle contextual information such as the traffic load \textcolor{orange}{\st{at the APs}}, which can \textcolor{orange}{be useful for} decision-making. Furthermore, mobility pattern prediction and predicted user requirements can be included in the system, thus empowering the association and handover mechanisms with insightful information. %\textcolor{orange}{Predictions can be provided, for instance, by Deep Neural Networks (DNN).}

@@

\subsubsection{Inference-Based Coordinated Scheduling}
Contrary to traditional cellular-type networks, WLAN deployments can be chaotic, especially in residential scenarios where anyone can set-up an AP and create a wireless network. This typically leads to complex scenarios \textcolor{orange}{where inter-BSS interactions prevent the existing scheduling approaches to ensuring quality of service.} \textcolor{orange}{Fortunately}, ML can be used to infer these interactions and provide a solution accordingly. In particular, through coordinated ML-assisted scheduling, different APs can trigger uplink/downlink transmissions from/to the appropriate stations (STAs), thus increasing the network throughput \textcolor{orange}{whilst} reducing the number of packet collisions.

\subsubsection{Reinforcement Learning-Based Spatial Reuse} 
\textcolor{orange}{\st{In contrast to centralized and cooperative ML approaches, local optimization is required for independent WLANs. In this regard,}} Spatial reuse aims to improve channel utilization by means of sensitivity adjustment mechanisms. However, selecting the best sensitivity threshold is not trivial, given the complex spatial interactions that occur among WLANs and the high action space. \textcolor{orange}{To that} purpose, reinforcement learning can be introduced to improve spectral efficiency in a decentralized manner, so that each WLAN learns based on experience.

@@

\begin{figure*}[!ht]
	\centering
	\includegraphics[width=0.5\textwidth]{itu_ml_architecture}
	\caption{ITU logical architecture for future networks \cite{itu2019architecture}. \textcolor{blue}{Entities contain input/output interfaces for communication, while the ML intent is a declarative file with information related to the use case.}}
	\label{fig:itu_ml_architecture}
\end{figure*}

@@

Figure \ref{fig:itu_ml_architecture} shows the elements that compose the ML overlay (management subsystem, multi-level ML pipeline, and closed-loop subsystem), which are further described in the following subsections. Based on this \textcolor{orange}{standard} overlay, ML applications can be instantiated in the logical entities (represented by white boxes).\textcolor{orange}{\st{, which are managed in a standard manner.}} %Communication among entities is done through input and output interfaces.

\subsection{Management Subsystem} 
The management subsystem is in charge of the deployment and the orchestration of the ML services \textcolor{orange}{that operate} in the underlying network. To that purpose, the \textcolor{blue}{machine learning function orchestrator (MLFO)} entity is defined. \textcolor{blue}{The MLFO is first instantiated by a declarative intent that uses a meta language, which specifies the ML use case to be applied, including initialization, policies and constraints. Then, the MLFO initializes the elements of the multi-level ML pipeline in the corresponding devices and monitors their operation during execution.}

\subsection{Multi-Level Machine Learning Pipeline} 
The multi-level ML pipeline is formed by a set of logical entities that perform the actual ML activity in a given network underlay. The ML pipeline is instantiated and managed by the management subsystem and it is mainly in charge of the data collection and pre-processing, ML model application, and output distribution. The logical entities that compose the ML pipeline are as follows:
\begin{itemize}
	\item \textbf{Source (src):} \textcolor{orange}{generates data to be used by the ML mechanism.}
	\item \textbf{Collector (C):} \textcolor{orange}{collects the data generated by sources.}
	\item \textbf{Pre-processor (PP):} \textcolor{orange}{prepares the data collected for its utilization by the ML mechanism.}
	\item \textbf{Model (M):} \textcolor{orange}{applies an ML model according to the use case.}
	\item \textbf{Policy (P):} \textcolor{orange}{provides a set of constraints and/or guidelines that delimit the behavior of the model.}
	\item \textbf{Distributor (D):} \textcolor{orange}{spreads the ML output across all the corresponding targets (or sinks).}
	\item \textbf{Sink (sink):} \textcolor{orange}{applies the ML output that is received from the distributor.}
\end{itemize}

\subsection{Closed-loop Subsystem} 
In order to address network dynamics, the ML operation is assisted by a closed-loop subsystem. In particular, a sandbox environment can be used to reproduce (internal network) or simulate (simulator/emulator) the effect of certain ML optimization, so that additional information is provided to the system beforehand. As for the ML pipeline, the closed-loop subsystem is orchestrated by the management subsystem. Network simulators (e.g., ns-3, Komondor \cite{barrachina2019komondor}) are examples of closed-loop subsystems, which can serve two purposes: \emph{i)} generate synthetic data to be used for training, and \emph{ii)} \textcolor{blue}{run simulations} for devising the potential of a given ML method to be applied afterward on the real network.

@@

IEEE 802.11 WLANs are versatile in the sense that many different kinds of deployments can be realized. Roughly, we differentiate between two main types:
\begin{itemize}
	\item \textbf{Enterprise:} a set of BSSs can be jointly operated from the edge and/or the cloud, thus providing management and orchestration functionalities (e.g., centralized authentication, channel allocation). Enterprise-like deployments are realized through extended service sets (ESS) and can be typically found in environments controlled by a single network operator, such as university campuses, offices, stadiums, etc. 
	\item \textbf{Residential:} each BSS is responsible for its own management and operation. In the context of residential scenarios (but not limited to), ad-hoc WLANs such as personal hotspots are gaining popularity for infrastructureless communications (i.e., peer-to-peer). Ad-hoc WLANs are realized through the independent basic service set (IBSS).
\end{itemize}

Figure \ref{fig:overview_learning_approaches} illustrates the enterprise and residential-like deployments, as well as a set of IEEE 802.11 mechanisms (summarized in Table \ref{tab:opportunities_amendments}) that can \textcolor{orange}{facilitate the adoption of the ML-based architecture in WLANs.} These mechanisms can be used for data collection, training, and output distribution.

@@

\textcolor{red}{TODO: convert table II into bullets.}

@@

%%% CHALLENGES IN IEEE 802.11 WLANs
\subsection{Challenges in IEEE 802.11 WLANs}
\label{section:ieee_80211_wlans}
The application of ML methods in WLANs is tightly tied to the technological challenges posed by these types of networks. The major challenges encountered in wireless communications stand for fast data expiry and lack of resources for data handling (e.g., storage, computation, information exchange). Regarding Wi-Fi networks, we find the following challenges:
\begin{enumerate}
	\item \textbf{Network dynamics:} channel fluctuations (e.g., due to multipath fading), users mobility and varying traffic needs entail a big challenge to ML applications. As a result of network dynamics, ML methods need to constantly adapt to the environment, which may be achieved through continuous re-training.
	\item \textbf{Limited communication resources:} since Wi-Fi operates under unlicensed bands, resources are scarce and shared. Thus, any potential communication required by a certain ML mechanism (as for distributed learning) may fail or be delayed if the medium is congested. As a result, the ML operation must be robust and resilient enough in order to react to potential communication issues.
	\item \textbf{Limited computation and storage resources:} computation and storage resources may also be limited in WLANs, especially in residential-like deployments. Therefore, the ML operation should include computation-efficient procedures. Another implication of limited resources lies in the availability of information to be used by ML algorithms, especially for online learning methods.
	\item \textbf{Adversarial environment:} in many cases, Wi-Fi deployments are chaotic in the sense that many overlapping \textcolor{orange}{BSSs} coexist without any kind of cooperation \textcolor{orange}{\st{(e.g., residential buildings)}}. This is a particularly interesting challenge for ML \textcolor{orange}{methods}, which may lead to adversarial settings where different agents compete for the same resources. Moreover, multi-vendor devices may implement different ML mechanisms that may lead to clashing interests.
	\item \textbf{Legacy devices:} \textcolor{orange}{similarly} to the previous case, ML-empowered WLANs may coexist with other legacy networks that do not perform any intelligent operation. It is then required for ML methods to be aware of those legacy devices, so that \textcolor{orange}{unfair} situations are avoided.
\end{enumerate}

Apart from the aforementioned WLAN-specific challenges, other inter-domain issues should be considered. First, security is required since ML mechanisms store and/or exchange sensitive data that may be exposed. Besides, interoperability should be tackled to allow the deployment of ML solutions to different underlay networks. In this regard, the standardized ITU ML pipeline stands up as a promising solution.
